The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2019-01-26 Paper.Rmd
edits I forgot to commit before

dt <- tibble(Description = c("1. P-value, independence test",
           "2. Minimum-AIC copula",
           "3. AIC weight of minimum-AIC copula",
           "4. AIC weight of N copula",
           "5. P-value, Cramer-von Mises goodness of fit test",
           "6. P-value, Kolmogorov-Smirnov goodness of fit test",
           "7. Model averaged lower-tail dependence",
           "8. Model averaged upper-tail dependence",
           "9. Model averaged lower- minus upper-tail dependence",
           "10. Asymmetry rank in surrogates")
)
tab_bivar_summary<-cbind(dt,tab_bivar_summary)
knitr::kable(tab_bivar_summary, 
             format="latex",
             caption = "Summary of Q1 results for bivariate datasets. The $p$-values in rows 5-6 are for the minimum-AIC copula. Model averaging used for rows 7-9 was based on AIC weights. For row 10, the statistic $P_{0,u_b}-P_{1-u_b}$ was computed for real data and for 1000 Kendall-preserving surrogate datasets with N copulas, and the rank of real-data value in the distribution of surrogate values is presented. If ranks were $>975$ or $<25$, asymmetry of tail dependence was significant compared to a N-copula null hypothesis. The value of $u_b$ used depended on the size of the dataset, and can be found in Figs. \\ref{SM-fig_soilCN_nonparam} to \\ref{SM-fig_cc2000_nonparam} and Tables \\ref{SM-tab_soilCN_asym}, \\ref{SM-tab_bmr_birds_asym}, \\ref{SM-tab_bmr_mammals_asym}, and \\ref{SM-tab_cc_asym_2000}. \\label{tab_bivar_summary}",
             booktabs=T, linesep = "\\addlinespace",
             col.names = c(" ", "Soil C and N", "Bird masses and BMR", "Mammal masses and BMR", "Cedar Creek data")
             )%>%column_spec(2:5,width="1.4cm")%>%
             column_spec(1, width = "8.5cm")%>%kable_styling(latex_options = "basic",position = "center")
```
<!--***DAN: Shya, it says Pl-Pu above, but I suspect it was P_first slice - P_last slice, right?
Can you please look back at the code, if necessary, and write a comment below this one 
indicating what results are actually presented in the table?
***Shya: Yes, you are right, here Pl-Pu means P_first slice - P_last slice. In table, item 10 : Assymetry rank is shown for Kendall preserving surroagates.
Reference : bivfunctionplot.R file, lines 134,135,161
***DAN: OK, great. I have made changes to deal with this, but am leaving the comment for now
as a reminder that we looked at this-->

@@

<!--Question 2 analyses-->
# Concepts and models for Q2: possible causes of non-normal copula structure \label{Causes}

This section is dedicated to exploring question \hyperref[Intro]{\textbf{(Q2)}} from \nameref{Intro}. We showed environmental copula structure and asymmetric sensitivities to the environment could produce population copula structures.
<!--We wanted to find how are copula structure of ecologically influential climate variables changing as part of climate change?--> 
<!--Moran effects-->
\textbf{(i)} We were interested to extend the basic idea of Moran's theorem : if environmental noises at two different habitat patches had some specific assymetric tail dependence then what will be the outcome dependence in the populations at those habitat patches. Should their dependence be exactly the same as of the noises or not ? To begin with, we chose a two habitat patch ($i = 1, 2$) with AR(1) model [Eq. \ref{eq.AR1}] where population $P_{i}$ at time $(t+1)$ depends upon its past value at $(t)$ and the iid noise (assumed to be caused by environmental fluctuations) $\epsilon_{i} (t)$.

@@

To keep $Var(P_{i}(t+1))$ equal to unity, we set noise coefficient $\alpha$ in terms of AR coefficient $\beta$ as $\alpha = \sqrt(1-\beta^2)$. We showed with Eq. \ref{eq.AR1} in the next section that both model selection method and non-parametric analysis indicated similar tail dependence in population copula as that of noise copula. <!--Asymmetric sensitivity and copula structure in populations-->
\textbf{(ii)} Secondly, we intended to see if any specific (lower or upper) tail dependence emerges from a copula of two meta-populations, due to some global environmental factors (like temperature or snowfall) which behaved differently beyond a certain expectation threshold (as shown in Eq. \ref{eq.AS_g}). To test the assymetric sensitivity, we constructed a two habitat patch model for population $P_{i} (i=1,2)$ as follows;

>>>>

dt <- tibble(Description = c("1. P-value, independence test",
           "2. Minimum-AIC copula",
           "3. AIC weight of minimum-AIC copula",
           "4. AIC weight of N copula",
           "5. P-value, Cramer-von Mises goodness of fit test",
           "6. P-value, Kolmogorov-Smirnov goodness of fit test",
           "7. Model averaged lower-tail dependence",
           "8. Model averaged upper-tail dependence",
           "9. Model averaged lower- minus upper-tail dependence",
           "10. Asymmetry rank in surrogates")
)
tab_bivar_summary<-cbind(dt,tab_bivar_summary)
knitr::kable(tab_bivar_summary, 
             format="latex",
             caption = "Summary of Q1 results for bivariate datasets. The $p$-values in rows 5-6 are for the minimum-AIC copula. Model averaging used for rows 7-9 was based on AIC weights. For row 10, the statistic $P_{0,0.1}-P_{0.9,1}$ was computed for real data and for 1000 Kendall-preserving surrogate datasets with N copulas, and the rank of real-data value in the distribution of surrogate values is presented. If ranks were $>975$ or $<25$, asymmetry of tail dependence was significant compared to a N-copula null hypothesis. The value of $u_b$ used depended on the size of the dataset, and can be found in Figs. \\ref{SM-fig_soilCN_nonparam} to \\ref{SM-fig_cc2000_nonparam} and Tables \\ref{SM-tab_soilCN_asym}, \\ref{SM-tab_bmr_birds_asym}, \\ref{SM-tab_bmr_mammals_asym}, and \\ref{SM-tab_cc_asym_2000}. \\label{tab_bivar_summary}",
             booktabs=T, linesep = "\\addlinespace",
             col.names = c(" ", "Soil C and N", "Bird masses and BMR", "Mammal masses and BMR", "Cedar Creek data")
             )%>%column_spec(2:5,width="1.4cm")%>%
             column_spec(1, width = "8.5cm")%>%kable_styling(latex_options = "basic",position = "center")
```
<!--***DAN: Shya, it says Pl-Pu above, but I suspect it was P_first slice - P_last slice, right?
Can you please look back at the code, if necessary, and write a comment below this one 
indicating what results are actually presented in the table?
***Shya: Yes, you are right, here Pl-Pu means P_first slice - P_last slice. In table, item 10 : Assymetry rank is shown for Kendall preserving surroagates.
Reference : bivfunctionplot.R file, lines 134,135,161
***DAN: OK, great. I have made changes to deal with this, but am leaving the comment for now
as a reminder that we looked at this-->

@@

<!--Question 2 analyses-->
# Background, concepts and methods for Q2: causes/mechanisms of non-normal copula structure \label{Causes}

Having demonstrated that non-N copula structure and asymmetric tail dependence are common
features of ecological data, we explore via mathematical 
models two possible mechanisms of this structure which 
seem likely to operate in population systems. The first mechanism 
is an extension of the well known Moran
effect, and was summarized in the Introduction. 
It explores the idea that non-N copula structure and asymmetric tail dependence 
between environmental variables measured in two places may bring about similarly 
structured dependence between ecological variables that are influenced by the 
environmental variables. Let $P_i(t)$ be a population index for location $i$ ($i=1,2$),
and assume the index follows AR(1) dynamics within a patch, 
$P_i(t+1)=\beta P_{i}(t-1)+ \sqrt{1-\beta^2} \epsilon_i(t)$ where the $\epsilon_i(t)$ are 
standard-normal random variables that are independent for distinct times, $t$, but correlated

To begin with, we chose a two habitat patch ($i = 1, 2$) with AR(1) model [Eq. \ref{eq.AR1}] where population $P_{i}$ at time $(t+1)$ depends upon its past value at $(t)$ and the iid noise (assumed to be caused by environmental fluctuations) $\epsilon_{i} (t)$.

@@

To keep $Var(P_{i}(t+1))$ equal to unity, we set noise coefficient $\alpha$ in terms of AR coefficient $\beta$ as $\alpha = \sqrt(1-\beta^2)$. We showed with Eq. \ref{eq.AR1} in the next section that both model selection method and non-parametric analysis indicated similar tail dependence in population copula as that of noise copula. <!--Asymmetric sensitivity and copula structure in populations-->
\textbf{(ii)}

The second mechanism, which was also summarized in the introduction,
relates to Liebig's law of the minimum, or, more generally, to the potential for 
environmental variables to influence ecological variables through nonlinear relationships.

Secondly, we intended to see if any specific (lower or upper) tail dependence emerges from a copula of two meta-populations, due to some global environmental factors (like temperature or snowfall) which behaved differently beyond a certain expectation threshold (as shown in Eq. \ref{eq.AS_g}). To test the assymetric sensitivity, we constructed a two habitat patch model for population $P_{i} (i=1,2)$ as follows;
